Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-02 manuscript/manuscript.tex

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb genomic windows which we
refer to as \emph{windowed covariances} (Figure \ref{fig:figure-3} shows these
distributions pooling across all replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance of each tile will be noisy, due to sampling and genetic drift, and
the neutral distribution of the covariance is complicated due to linkage
disequilibria (which can occur over long physical distances in E\&R and
selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To address
this, we have developed a permutation-based procedure that constructs a null
distribution by randomly flipping the signs of the  allele frequency changes
per-genomic window. This destroys the systematic covariances created by linked
selection and creates a sampling distribution of the covariances spuriously
created by neutral genetic drift while preserving the complex dependencies
between adjacent loci created by linkage disequilibrium. This empirical neutral
null distribution is conservative in the sense that the variances of the
covariances are wider than expected under drift alone as they include the
effect of selection on the allele frequency change within a time-interval, just
not between time-intervals. We see (Figure \ref{fig:figure-3} A and B) that
windowed temporal covariances between close timepoints are skewed positive (a
heavy right tail), while between more distant timepoints these windowed
temporal covariances tend to shift to become more negative (a heavy left tail).
We quantified the degree to which the left and right tails are inflated
compared to the null distribution as a function of time, and see excesses in
both tails in Figure \ref{fig:figure-3} C. This finding is also robust to
sign-permuting allele frequency changes on a chromosome-level, the longest
extent that gametic linkage disequilibria can extend (Supplementary Figure
\ref{suppfig:barghi-tailprobs-seqid}). We see a striking pattern that the
windowed covariances not only decay towards zero, but in fact become negative
through time, consistent with many regions in the genome having had a reversed
fitness effect at later timepoints.

% (see
% Supplementary Material Section \ref{supp:empirical-null} for more details on
% this approach).

@@

%In our reanalysis of the
%\textcite{Castro2019-uk} data, by showing the covariance signal remains even
%after excluding the chromosomes containing two large-effect loci, we showed the
%covariance signal we detect is indeed driven by polygenic fitness variation
%(and confirm a point in the original study that tibiae length, aside from these
%two large effect loci, is a highly polygenic trait).

%This work has led
%to the discovery of selected genes from genomic signals of hitchhiking
%\parencite{Nair2003-tw,Voight2006-rn}, found genome-wide signals of linked
%selection
%\parencite{Aguade1989-jx,Begun1992-ey,Cutter2010-gi,Andersen2012-bj,Cutter2003-tl}
%leading to the development of recurrent hitchhiking \parencite{Stephan1992-jc}
%and background selection \parencite{Charlesworth1993-gb} models that attempt
%explain this pattern. Since, much attention has focused on quantifying the
%impact of background selection \parencite{McVicker2009-ax} and developing
% theoretic machinery \parencite{Coop2012-cd} to differentiate the regional
% reductions in diversity into those caused by full and partial sweeps, and those
% caused by background selection \parencite{Elyashiv2016-vt}. However, as other
% models of soft sweeps from standing variation emerged
% \parencite{Hermisson2005-hs,Pennings2006-lj}, attention shifted towards
% detecting other modes of linked selection \parencite{Pritchard2010-tk}, a more
% difficult endeavor \parencite{Przeworski2005-bg} addressed using new coalescent
% models \parencite{Berg2015-xj} and machine learning methods
%\parencite{Schrider2017-yx}.

@@

While it challenging to apply temporal methods to natural populations
there is a lot of promise for these approaches \cite{Dmitri_others}. 
Efforts to estimate the quantify the impact of linked selection in
humans and {\it Drosophila melanogaster} have suggested that XXX to
XXX of allele frequency change over long time periods is driven by
selection \citep{}. Thus there is reason to be hopeful that we could
detect this from temporal data, which would help to resolve the
time-scales that linked selection act over. In our reanalysis of the \textcite{Barghi2019-qy} study, we find evidence
of complex linked selection dynamics, with selection pressures
flipping over time due to either environmental change, the breakup of
epistatic combinations or advantagous haplotypes. Such patterns would
 be completely obscured in samples only from contempory populations,
 thus we can hope tyo have a much richer picture of linked selection
 as temporal sequencing becomes more common.

%Overall, we hope this will encourage more temporal genomic studies in both
%laboratory and natural populations.

@@

% \subsection{The Empirical Neutral Null Windowed Covariance Distribution}
% \label{supp:empirical-null}

% XXX

@@

  \caption{The 20\% lower and upper tail probabilities for the observed
  windowed covariances from the \textcite{Barghi2019-qy} study, based on
  sign-permuting at the chromosome level. This permutation empirical null
  is robust to long-range linkage disequilibrium acting over entire chromosomes.}

>>>>

Finally, we observed that in the longest study  we analyzed
\parencite{Barghi2019-qy}, some genome-wide temporal covariances become
negative at future timepoints (see the first two rows in Figure
\ref{fig:figure-1} A). This shows that alleles that were on average going up
initially are later going down in frequency, i.e. that the average direction of
selection experienced by alleles has flipped. This must reflect either a change
in the environment or the genetic background, due to epistatic relationships
among alleles altered by frequency changes or recombination breaking up
selective alleles.  Such reversals of selective dynamics could be occurring at
other timepoints but the signal of a change in the direction of selection at
particular loci may be washed out when we calculate our genome-wide average
temporal covariances. To address this limitation, we calculated the
distribution of the temporal covariances over 100kb genomic windows which we
refer to as \emph{windowed covariances} (Figure \ref{fig:figure-3} shows these
distributions pooling across all replicates; see Supplementary Figure
\ref{suppfig:barghi-offset-replicate-panels} for individuals replicates). The
covariance of each tile will be noisy, due to sampling and genetic drift, and
the neutral distribution of the covariance is complicated due to linkage
disequilibria (which can occur over long physical distances in E\&R and
selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}). To address
this, we have developed a permutation-based procedure that constructs an
empirical null distribution by randomly flipping the signs of the allele
frequency changes per-genomic window. This destroys the systematic covariances
created by linked selection and creates a sampling distribution of the
covariances spuriously created by neutral genetic drift while preserving the
complex dependencies between adjacent loci created by linkage disequilibrium.
This empirical neutral null distribution is conservative in the sense that the
variances of the covariances are wider than expected under drift alone as they
include the effect of selection on the allele frequency change within a
time-interval, just not between time-intervals. We see (Figure
\ref{fig:figure-3} A and B) that windowed temporal covariances between close
timepoints are skewed positive (a heavy right tail), while between more distant
timepoints these windowed temporal covariances tend to shift to become more
negative (a heavy left tail).  We quantified the degree to which the left and
right tails are inflated compared to the null distribution as a function of
time, and see excesses in both tails in Figure \ref{fig:figure-3} C. This
finding is also robust to sign-permuting allele frequency changes on a
chromosome-level, the longest extent that gametic linkage disequilibria can
extend (Supplementary Figure \ref{suppfig:barghi-tailprobs-seqid}). We see a
striking pattern that the windowed covariances not only decay towards zero, but
in fact become negative through time, consistent with many regions in the
genome having had a reversed fitness effect at later timepoints.

@@

@@

While it challenging to apply temporal methods to natural populations there is
a lot of promise for these approaches
\parencite{Bergland2014-ij,Machado2018-cs}. Efforts to estimate and quantify
the impact of linked selection in humans and \emph{Drosophila melanogaster}
have suggested that XXX to XXX of allele frequency change over long time
periods is driven by selection \parencite{}. Thus there is reason to be hopeful
that we could detect this from temporal data, which would help to resolve the
timescales that linked selection act over. In our reanalysis of the
\textcite{Barghi2019-qy} study, we find evidence of complex linked selection
dynamics, with selection pressures flipping over time due to either
environmental change, the breakup of epistatic combinations or advantageous
haplotypes. Such patterns would be completely obscured in samples only from
contemporary populations, thus we can hope to have a much richer picture of
linked selection as temporal sequencing becomes more common.

@@

\subsection{The Empirical Neutral Null Windowed Covariance Distribution}
\label{supp:empirical-null}

\vb{To detect an excess of genomic regions with unusually high or low covariances,
we need to compare the distribution of observed windowed covariances to a null
distribution of windowed covariances that we would expect under no selection.
While we could construct a theoretic sampling distribution of the spurious
covariances created by neutral genetic drift at particular site, the unknown
linkage disequilibrium between sites would mean that this is not an adequate
null model for the distribution of windowed covariances in our data.

To address this limitation, we construct a neutral null model by sign-permuting
the observed allele frequency changes. This destroys the covariances built up
by selection, mimicking a neutral allele's frequency trajectory. This approach
is conservative, since selection also acts to increase the magnitude of allele
frequency changes (see equation 1 of \cite{Buffalo2019-io}), but this magnitude
is not affected by the sign-permutation procedure. Consequently, the resulting
empirical null distribution has higher variance than would be expected under
neutrality alone.

Since variants have unknown dependency structure due to LD, we sign-permute at
the window level (analogous to our block bootstrap approach). Our covariance
statistic, $Q_{t,s} = \nicefrac{1}{L} \sum_l \Delta p_{t,l} \Delta p_{t,l}$
(for simplicity, we drop the replicate index for clarity, do not center
observations since their expected change is zero, and assume $|t-s| > 1$ to
avoid the need for a bias correction) has a variance that depends on the LD:

\begin{align}
  \var(Q_{t,s}) &= \frac{1}{L^2} \sum_l \var(\Delta p_{t,l} \Delta p_{s,l}) + \frac{1}{L^2} \sum_{i \ne j} \cov(\Delta p_{t,i}\Delta p_{s,i}, \Delta p_{t,j}\Delta p_{s,j}) 
\end{align}

Under neutrality alone, we can simplify this, using the fact that allele
frequency changes are independent through time,

\begin{align}
  \var(Q_{t,s}) &= \frac{1}{L^2} \sum_l \E(\Delta p_{t,l}^2) \E(\Delta p_{s,l}^2) + \frac{1}{L^2} \sum_{i \ne j} \E(\Delta p_{t,i} \Delta p_{s,i} \Delta p_{t,j} \Delta p_{t,j}) \\
  &= \frac{1}{L^2} \sum_l \frac{p_{t,l}(1-p_{t,l}) p_{s,l}(1-p_{s,l})}{4N^2} + \\ \nonumber 
  & \;\;\;\;\;\; \frac{1}{L^2} \sum_{i \ne j} \frac{\left(p_{s,i} p_{s,j}-4 D^2 \left(1-2 N r^2\right) \right) \left( p_{t,i} p_{t,j}-4 D^2 \left(1-2 N r^2\right) \right)}{4 N^2}
\end{align}
%

where this last line can be found by assuming haplotype frequencies are sampled
according to a Multinomial process, and recombination occurs after sampling
(see Barton and Otto, 2005). This indicates that the variance of the covariance
statistic, averaged over loci, depends on the linkage disequilibrium term $D^2$
even without the presence of linked selection. However, under neutrality alone,
the covariance statistic is unaffected by LD with sites outside averaged over.
Still, we sign-permute loci at the chromosome level and find that the tail
probabilities do still indicate a reverse in the direction of selection. }

@@

  \caption{The 20\% lower and upper tail probabilities for the observed
    windowed covariances from the \textcite{Barghi2019-qy} study, based on
    sign-permuting at the chromosome level. This permutation empirical null is
    robust to long-range linkage disequilibrium acting over entire chromosomes
  (see Supplementary Material section \ref{supp:empirical-null}).}
